---
title: "Data preparation of SF Cycle Share data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, echo=TRUE, warning=FALSE, results='hide', message=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
library(tidyr)
```

## Introduction

Dataset source: https://www.kaggle.com/benhamner/sf-bay-area-bike-share/version/1

I haven't really analyzed San Francisco Cycle Share data yet. It should make for a good blog post or two (or Kaggle kernel) as well as an EDA assignment in R for pcda class. There are several files:

- station.csv - one row per station. Includes lat/lon, city, dock count and install date
- status.csv - like a hillmaker bydate table at the minute level by station with bikes available and docks available
- trip.csv - one row per trip with standard fields. Includes a zip code that is user entered (according to answer in Discussion forum)
- weather.csv - one row per date per zip code with min, mean, and max of several weather related variables. In Discussion forum, I found this about the zip code:

    Also, FYI: There are five cities that the BABS program operates in including the following:

    city <- c ("San Francisco", "Redwood City", "Palo Alto", "Mountain View", "San Jose")
    
    These should correspond with…
    
    zip_code
    [1] 94107 94063 94301 94041 95113

In addition there's a SQLite database containing the same four data tables. 

**TODO** - check if DB Browser for SQLite is installed on pcda vm.


## Station data

Read in the csv file and do a little data type fixing using the readr package.

```{r read_data}
station <- read_csv("data/raw/station.csv",
    col_types = cols(id = col_integer(),
                     dock_count = col_integer(),
                     installation_date = col_date(format = "%m/%d/%Y")))

station$id <- as.factor(station$id)
str(station, give.attr=FALSE)
```

Hmmm, should we convert `name` and `city` to factors? The `name` values will be in the `trip` table and
we'll want to use join type operations to get station level information (e.g. the lat/lon). I guess
I never thought carefully about this and was just trusting base R `merge` and dplyr joins to coerce
factors to strings if needed (e.g. the levels in the two tables don't match perfectly). Some
good posts on this are:

- https://rpubs.com/hadley/45044
- https://stackoverflow.com/questions/23802571/safely-merge-data-frames-by-factor-columns
- https://stackoverflow.com/questions/30468412/dplyr-join-warning-joining-factors-with-different-levels

For dplyr, coercion to strings is done if needed and a warning raised.

## Trip data

This is the primary table we'll use for analysis.

```{r readtrip, warning=FALSE}
trip <- read_csv("data/raw/trip.csv",
                 col_types = cols(id = col_integer(),
                     start_station_id = col_integer(),
                     end_station_id = col_integer(),
                     bike_id = col_integer(),
                     zip_code = col_integer()))

```

Convert start and end dates to date times.

```{r convdates}
trip$start_date <- as.POSIXct(trip$start_date, format="%m/%d/%Y %H:%M")
trip$end_date <- as.POSIXct(trip$end_date, format="%m/%d/%Y %H:%M")
```

Let's convert some of the fields to factors.

```{r factors}
trip$start_station_id <- as.factor(trip$start_station_id)
trip$end_station_id <- as.factor(trip$end_station_id)
trip$subscription_type <- as.factor(trip$subscription_type)
trip$zip_code <- as.factor(trip$zip_code)
```

For analytical convenience, let's create some new fields in the `trip` data frame.

### Date parts

Use the `start_date` as a basis and add some date part fields.

```{r feng_datepart}
trip <- trip %>% 
  mutate(trip_date = date(start_date),
         start_year = year(start_date),
         start_month = month(start_date),
         start_hour = hour(start_date),
         start_dow = wday(start_date),
         trip_ym = floor_date(start_date, "month"))
```

We'll likely want to do some day of week analysis.

```{r feng_dow}
trip$start_dow <- as.factor(trip$start_dow)
levels(trip$start_dow) <- c("Su", "Mo", "Tu", "We", "Th", "Fr", "Sa")

trip$start_weekday <- trip$start_dow %in% 2:6
trip$start_weekday <- as.factor(trip$start_weekday)
levels(trip$start_weekday) <- c("Weekend", "Weekday")
```

### Trip type

If start and end points are the same, it's a "round trip".

```{r round_trip}
trip <- trip %>% 
  mutate(round_trip = (start_station_id == end_station_id))

trip$round_trip <- as.factor(trip$round_trip)
levels(trip$round_trip) <- c("One way", "Round trip")
```


### Trip geography

Let's pull in the start and end cities (based on zip code in the `station` data frame) and create a few more fields based on the starting location and on ending location.

```{r city1}
trip <- trip %>% 
  left_join(station, by = c("start_station_id" = "id")) %>%
  mutate(start_city = city,
         start_lat = lat,
         start_long = long) %>% 
  select(-c(city, name, lat, long, dock_count, installation_date))

```

```{r city2}
trip <- trip %>% 
  left_join(station, by = c("end_station_id" = "id")) %>%
  mutate(end_city = city,
         end_lat = lat,
         end_long = long) %>% 
  select(-c(city, name, lat, long, dock_count, installation_date))

```

Here's our prepped `trip` data frame:

```{r trip_final}
str(trip)
```

## Weather data

```{r readweather}
weather <- read_csv("data/raw/weather.csv")
str(weather)
summary(weather)
```

Fix up data types for `date` and `precipitation_inches`.

```{r}
weather$date <- as.POSIXct(weather$date, format="%m/%d/%Y")
weather$precipitation_inches <- as.numeric(weather$precipitation_inches)
```

```{r}
summary(weather)
```

## Save data

```{r}
save(station, trip, weather, file = "data/processed/sfcs_prepped.Rdata")
```

