---
title: "EDA for SFCS - station balance"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(tidyr)
library(readr)
```

## Introduction

Let's look at station balance

- net source or net destination
- daily balance
- number of bikes in use
- number of bikes at station



## Load data

In initial_exploration_sfcs.Rmd we did some data prep work and saved updated versions of the data.

```{r load_data}
load("sfcs_prepped.Rdata")
```

```{r read_status}
status <- read_csv("data/raw/status.csv")
str(status)
summary(status)
```

## Ins and outs

Let's compute number of trips per day (?) out and in by station. Are inventories balanced at night? Status table should reveal the answer.

```{r}
trip %>% 
  group_by(start_station_id, trip_date) %>% 
  count()
```

```{r}
startdate <- ymd("2014-01-01")
enddate <- ymd("2015-08-31")

dates <- seq(startdate, enddate, by='days')
stations <- station$id
stations2 <- unique(trip$start_station_id)
stations3 <- unique(trip[,"start_station_id"])
date_station <- expand.grid(stations, dates)
names(date_station) <- c("start_station_id", "trip_date")

# Get sorted by station and date
date_station <- date_station %>% 
  arrange(start_station_id, trip_date)

# Count trips out
num_out_station_date <- trip %>% 
  filter(between(trip_date, startdate, enddate)) %>% 
  group_by(start_station_id, trip_date) %>% 
  count()

# Join the fully seeded table, `date_hour_sub`, to `tot_trips_bydatehour_sub`. 

num_out_station_date <- date_station %>% 
  left_join(num_out_station_date, by = c("start_station_id", "trip_date"))

# Replace the NAs with 0's.
num_out_station_date$n <- num_out_station_date$n %>% replace_na(0)


```