---
title: "EDA for SFCS - station balance"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(tidyr)
library(readr)
```

```{r}
source("bikeshare_util.R")
```


## Introduction

Let's look at station balance

- net source or net destination
- daily balance
- number of bikes in use
- number of bikes at station



## Load data

In initial_exploration_sfcs.Rmd we did some data prep work and saved updated versions of the data.

```{r load_data}
load("sfcs_prepped.Rdata")
```

```{r read_status}
# status <- read_csv("data/raw/status.csv")
# str(status)
# summary(status)
```

## Ins and outs

Let's compute number of trips per day (?) out and in by station. Are inventories balanced at night? Status table should reveal the answer.

```{r}
trip %>% 
  group_by(start_station_id, trip_date) %>% 
  count()
```

Do it manually first and then convert to function.

```{r}
startdate <- ymd("2014-01-01")
enddate <- ymd("2015-08-31")

# Get unique values of date and grouping field
dates <- seq(startdate, enddate, by='days')
stations <- station$id

# Use expand.grid to create all the combinations (the fully seeded table)
date_station <- expand.grid(stations, dates)
names(date_station) <- c("start_station_id", "trip_date")

# Get sorted by station and date
date_station <- date_station %>% 
  arrange(start_station_id, trip_date)

# Count trips out
num_out_station_date <- trip %>% 
  filter(between(trip_date, startdate, enddate)) %>% 
  group_by(start_station_id, trip_date) %>% 
  count()

# Join the fully seeded table, `date_station`, to `num_out_station_date`. 

num_out_station_date <- date_station %>% 
  left_join(num_out_station_date, by = c("start_station_id", "trip_date"))

# Replace the NAs with 0's.
num_out_station_date$n <- num_out_station_date$n %>% replace_na(0)


```

This is such a common task, it would be nice to have a function to do it. Ideally,
it would be nice to pass multiple grouping fields along with a date range.

```{r}
num_out <- counts_by_date_group(trip, startdate, enddate, trip_date, start_station_id)
num_out
```

```{r}
num_in <- counts_by_date_group(trip, startdate, enddate, trip_date, end_station_id)
num_in
```

Compute overall flow balance as `num_in - num_out`.

```{r}
flow_balance_date <- num_in %>% 
  left_join(num_out, by = c("end_station_id" = "start_station_id", "trip_date"))

names(flow_balance_date) <- c("station_id", "trip_date", "num_in", "num_out")

flow_balance_date <- flow_balance_date %>% mutate(
  balance = num_in - num_out
)

flow_balance_date
```

Obviously for round trips, in and out balance for each trip. It would be nice to do this
with `round_trip` as an additional grouping field.

```{r}
num_out_type <- counts_by_date_groups(trip, startdate, enddate, 
                                   trip_date, start_station_id, round_trip)

num_in_type <- counts_by_date_groups(trip, startdate, enddate, 
                                   trip_date, end_station_id, round_trip)

flow_balance_type_date <- num_in_type %>% 
  left_join(num_out_type, 
            by = c("end_station_id" = "start_station_id", "round_trip", "trip_date"))

names(flow_balance_type_date) <- c("station_id", "trip_type", "trip_date", "num_in", "num_out")

flow_balance_type_date <- flow_balance_type_date %>% mutate(
  balance = num_in - num_out
)

flow_balance_type_date
```

```{r}
flow_balance_type_date %>% 
  filter(trip_type == "Round trip")
```

```{r}
flow_balance_type_date %>% 
  filter(trip_type == "Round trip") %>% 
  summarize(
    tot_abs_balance = sum(abs(balance))
  )
```


```{r}
flow_balance_type_date %>% 
  filter(round_trip == "Round trip", abs(balance) > 0)
```

```{r}
table(trip$start_station_id)
```

```{r}
flow_balance_type_date %>% 
  filter(trip_type == "One way", station_id == 60) %>% 
  ggplot() + geom_histogram(aes(x = balance))
  
```

```{r box_station}

ggplot(flow_balance_date) + geom_boxplot(aes(x = as.factor(station_id), y = balance))
  
```

Let's zoom in.

```{r box_station2}

ggplot(flow_balance_date) + 
  geom_boxplot(aes(x = as.factor(station_id), y = balance)) +
  coord_cartesian(ylim = c(-50, 50))
  
```

```{r box_station3}

ggplot(flow_balance_date) + 
  geom_boxplot(aes(x = as.factor(station_id), y = balance)) +
  coord_cartesian(ylim = c(-30, 30))
  
```

```{r}
flow_balance_date %>% 
  group_by(station_id) %>% 
  summarize(
    mean_balance = mean(balance),
    sd_balance = sd(balance),
    cv_balance = sd_balance / mean_balance,
    p01_balance = quantile(balance, 0.01),
    p05_balance = quantile(balance, 0.05),
    q1_balance = quantile(balance, 0.25),
    median_balance = median(balance),
    q3_balance = quantile(balance, 0.75),
    p95_balance = quantile(balance, 0.95),
    p99_balance = quantile(balance, 0.99)
  ) %>% 
  arrange(mean_balance)
```

