---
title: "Initial exploration of SF Cycle Share data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
```

## Introduction

Dataset source: https://www.kaggle.com/benhamner/sf-bay-area-bike-share/version/1

I haven't really analyzed San Francisco Cycle Share data yet. It should make for a good blog post or two (or Kaggle kernel) as well as an EDA assignment in R for pcda class. There are several files:

- station.csv - one row per station. Includes lat/lon, city, dock count and install date
- status.csv - like a hillmaker bydate table at the minute level by station with bikes available and docks available
- trip.csv - one row per trip with standard fields. Includes a zip code that is user entered (according to answer in Discussion forum)
- weather.csv - one row per date per zip code with min, mean, and max of several weather related variables. In Discussion forum, I found this about the zip code:

    Also, FYI: There are five cities that the BABS program operates in including the following:

    city <- c ("San Francisco", "Redwood City", "Palo Alto", "Mountain View", "San Jose")
    
    These should correspond with…
    
    zip_code
    [1] 94107 94063 94301 94041 95113

In addition there's a SQLite database containing the same four data tables. 

**TODO** - check if DB Browser for SQLite is installed on pcda vm.


## Read in and first look at the data tables

### The station table

I'm going to read station.csv in a few different ways.

```{r read_station}
# Base with defaults (stringsAsFactors = TRUE)
station0 <- read.csv("~/Documents/projects/sf_cycle_share/data/raw/station.csv")

# readr with defaults
station1_readr <- read_csv("data/raw/station.csv")

# readr and fix a few column types
station2_readr <- read_csv("data/raw/station.csv",
    col_types = cols(dock_count = col_integer(),
                     installation_date = col_date(format = "%m/%d/%Y")))

```

Using base, the station name and city are treated as factors (correctly) but the date is also
treated as a factor (since R guessed it was a string) and we want it to be a date.

```{r}
str(station0)
```

Can we convert installation_date to a POSIXct value even though it's a factor?

```{r}
station0$installation_date <- as.POSIXct(station0$installation_date, format = "%m/%d/%Y")
str(station0)
```

Yep, that works.

With readr and using defaults, it's like using base with `stringsAsFactors = FALSE`.

```{r}
str(station1_readr)
```

With readr and fixing some data types, we can get what we want.

```{r}
str(station2_readr)
```

We'll use `station2_readr` and get rid of the others.

```{r}
station <- station2_readr
rm(station0, station1_readr, station2_readr)
```

```{r}
summary(station)
```

Should we convert `name` and `city` to factors? The `name` column will be in the `trip` table and
we'll want to use join type operations to get station level information (e.g. the lat/lon). I guess
I never thought carefully about this and was just trusting base R `merge` and `dplyr` to coerce
factors to strings if needed (e.g. the levels in the two tables don't match perfectly). Some
good posts on this are:

- https://rpubs.com/hadley/45044
- https://stackoverflow.com/questions/23802571/safely-merge-data-frames-by-factor-columns
- https://stackoverflow.com/questions/30468412/dplyr-join-warning-joining-factors-with-different-levels

For dplyr, coercion to strings is done if needed and a warning raised.

### Trip table

This is the primary table we'll use for analysis.

```{r readtrip}
trip <- read_csv("data/raw/trip.csv")
str(trip)
summary(trip)
```

Let's confirm we didn't lose any rows with those parsing errors.

```{bash}
wc -l data/raw/trip.csv
```

```{r}
head(trip)
```

Convert start and end dates to dates.

```{r}
trip$start_date <- as.POSIXct(trip$start_date, format="%m/%d/%Y %H:%M")
trip$end_date <- as.POSIXct(trip$end_date, format="%m/%d/%Y %H:%M")
str(trip)
```

```{r}
head(trip)
```

### Weather data

```{r readweather}
weather <- read_csv("data/raw/weather.csv")
str(weather)
summary(weather)
```

Fix up data types for `date` and `precipitation_inches`.

```{r}
weather$date <- as.POSIXct(weather$date, format="%m/%d/%Y")
weather$precipitation_inches <- as.numeric(weather$precipitation_inches)
```

```{r}
summary(weather)
```

## Preliminary exploration of trip data

```{r}
summary(trip)
```

Hmm, that max duration looks a little suspect.

```{r}
trip %>% 
  arrange(desc(duration))
```


For now, let's leave all the records intact.


Let's convert some of the fields to factors.

```{r}
trip$subscription_type <- as.factor(trip$subscription_type)
trip$zip_code <- as.factor(trip$zip_code)
```

Use `table` function to get trip counts by factor levels.

```{r}
table(trip$subscription_type)
```

```{r}
table(trip$start_station_name)
```

```{r}
head(trip)
```

## Feature engineering

For analytical convenience, let's create some new fields in the `trip` dataframe.

### Date parts

Use the `start_date` as a basis.

```{r}
trip %>% 
  mutate(trip_date = date(start_date),
         start_year = year(start_date),
         start_month = month(start_date),
         start_hour = hour(start_date),
         start_dow = wday(start_date)) %>% 
  select(starts_with("start_")) %>% 
  head(15)
```

Create the new fields.

```{r}
trip <- trip %>% 
  mutate(trip_date = date(start_date),
         start_year = year(start_date),
         start_month = month(start_date),
         start_hour = hour(start_date),
         start_dow = wday(start_date),
         trip_ym = floor_date(start_date, "month"))
```

```{r}
trip$start_weekday <- trip$start_dow %in% 2:6
trip$start_dow <- as.factor(trip$start_dow)
levels(trip$start_dow) <- c("Su", "Mo", "Tu", "We", "Th", "Fr", "Sa")


trip$start_weekday <- as.factor(trip$start_weekday)
levels(trip$start_weekday) <- c("Weekend", "Weekday")
```


### Trip type

If start and end points are the same, it's a "round trip".

```{r}
trip %>% 
  mutate(round_trip = (start_station_id == end_station_id)) %>% 
  select(start_station_id, end_station_id, round_trip) %>% 
  head(15)
```

```{r}
trip <- trip %>% 
  mutate(round_trip = (start_station_id == end_station_id))
```

### Trip geography

Let's pull in the start and end cities (based on zip code in the `station` dataframe).

```{r}
trip %>% 
  left_join(station, by = c("start_station_id" = "id")) %>% 
  select(contains("id"), city)
```


```{r}
trip %>% 
  left_join(station, by = c("start_station_id" = "id")) %>%
  mutate(start_city = city) %>% 
  select(contains("id"), contains("city"))
```

Let's create a few more fields based on the starting location and on ending location.

```{r}
trip <- trip %>% 
  left_join(station, by = c("start_station_id" = "id")) %>%
  mutate(start_city = city,
         start_lat = lat,
         start_long = long) %>% 
  select(-c(city, name, lat, long, dock_count, installation_date))

```

```{r}
trip <- trip %>% 
  left_join(station, by = c("end_station_id" = "id")) %>%
  mutate(end_city = city,
         end_lat = lat,
         end_long = long) %>% 
  select(-c(city, name, lat, long, dock_count, installation_date))

```

Let's save the dataframes.

```{r}
save(trip, station, weather, file="sfcs_prepped.Rdata")
```

## EDA of trips

I'll mostly use dplyr but will thrown in some apply type functions for practice.

### What are popular rental locations and trips?

Popular start locations

```{r}
trip %>% 
  group_by(start_station_name) %>% 
  summarize(
    num_trips = n()
  )
```



```{r count_starting}
trip %>% 
  group_by(start_station_name) %>% 
  summarize(
    num_trips = n()
  ) %>% 
  arrange(desc(num_trips))
```

Popular ending locations

```{r count_ending}
trip %>% 
  group_by(end_station_name) %>% 
  summarize(
    num_trips = n()
  ) %>% 
  arrange(desc(num_trips))
```

Popular trips

```{r count_trips}
trip %>% 
  group_by(start_station_id, end_station_id) %>% 
  summarize(
    num_trips = n()
  ) %>% 
  arrange(desc(num_trips))
```

Popular round trips

```{r count_roundtrips}
trip %>% 
  filter(round_trip) %>% 
  group_by(start_station_name) %>% 
  summarize(
    num_trips = n()
  ) %>% 
  arrange(desc(num_trips))
```

### What percent of trips are round trips?

Let's use ta
```{r}
table(trip$round_trip)
```

```{r}
prop.table(x = table(trip$round_trip))
```

Try to do it with dplyr.



This feels like it should work but doesn't.
```{r}
trip %>% 
  group_by(round_trip) %>% 
  summarize(
    pct = n() / sum(n())
  )
```

The dplyr `tally` command looks like the trick.

https://dplyr.tidyverse.org/reference/tally.html

```{r}
trip %>% 
  group_by(round_trip) %>% 
  tally()
```

```{r}
trip %>% 
  group_by(round_trip) %>% 
  tally() %>% 
  mutate(pct = n / sum(n))
```

Let's try with multiple grouping fields. Will the percentages be within groups?

```{r}
trip %>% 
  group_by(start_city, round_trip) %>% 
  tally() %>% 
  mutate(pct = n / sum(n))
```

```{r}
trip %>% 
  group_by(start_city, round_trip) %>% 
  tally()
```

```{r}
trip %>% 
  group_by(start_city, round_trip) %>% 
  tally() %>% 
  tally()
```

To get individual cell frequencies, need to divide by total number of rows in `trip` dataframe.

```{r cellfreq}
trip %>% 
  group_by(start_city, round_trip) %>% 
  tally() %>% 
  mutate(pct = n / nrow(trip))
```

Let's do city to city counts.

```{r}
table(trip$start_city, trip$end_city)
```

Obviously, the vast majority of trips are within San Francisco.

What exactly is the `table()` function returning.

```{r}
class(table(trip$start_city, trip$end_city))
```

According to the doc page at https://www.rdocumentation.org/packages/base/versions/3.5.2/topics/table, a
`table` is essentially an array of integers.

```{r}
prop.table(table(trip$start_city, trip$end_city))
```

### What is proporation of subscribers?

```{r}
trip %>% 
  group_by(subscription_type) %>% 
  count() %>% 
  mutate(pct = n / nrow(trip))
```

```{r}
trip %>% 
  group_by(subscription_type, round_trip) %>% 
  tally() %>% 
  mutate(pct = n / sum(n))
```